Classical mechanics as the high-entropy limit \\ of quantum mechanics

### 2024-12-03 Articles/2024-ClassicalLimit/Submissions/SciPost Physics/classicalLimitArticleSciPostPhys.tex
Updated arxiv version and submission version to SciPost Physics

\documentclass{article}
\usepackage{authblk}

% Language setting
% Replace `english' with e.g. `spanish' to change the document language
\usepackage[english]{babel}
\usepackage{ulem}
% Set page size and margins
% Replace `letterpaper' with `a4paper' for UK/EU standard size
\usepackage[letterpaper,top=2cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}

% Useful packages
\usepackage{assumptionsofphysics}
\usepackage[dvipsnames]{xcolor}
\usepackage{tikz-cd}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{pgfplots}

\def\>{\rangle}
\def\<{\langle}

\title{Classical mechanics as the high-entropy limit \\ of quantum mechanics}
\author[1]{Gabriele Carcassi}
\author[2]{Manuele Landini}
\author[1]{Christine A. Aidala}

\affil[1]{Physics Department, University of Michigan, Ann Arbor, Michigan 48109, USA}

\affil[2]{Institut f\"ur Experimentalphysik und Zentrum f\"ur Quantenphysik, \protect\\ Universit\"at Innsbruck, Innsbruck, Austria}

\begin{document}
\maketitle

\begin{abstract}
We show that classical mechanics can be recovered as the high-entropy limit of quantum mechanics. That is, the high entropy masks quantum effects, and mixed states of high enough entropy can be approximated with classical distributions. The mathematical limit $\hbar \to 0$ can be reinterpreted as setting the zero entropy of pure states to $-\infty$, in the same way that non-relativistic mechanics can be recovered mathematically with $c \to \infty$. Physically, these limits are more appropriately defined as $S \gg 0$ and $v \ll c$. Both limits can then be understood as approximations independently of what circumstances allow those approximations to be valid. Consequently, the limit presented is independent of possible underlying mechanisms and of what interpretation is chosen for both quantum states and entropy. 
\end{abstract}

\section{Introduction}

@@

Mathematically, it is common to see both limits as group contractions. This too can be given a physically meaningful interpretation. The low-speed limit is mathematically recovered by taking the highest possible speed and setting it to infinity. This way, all finite velocities are small compared to the highest possible speed. This corresponds to taking $c \to \infty$. Conversely, the high-entropy limit is mathematically recovered by taking the lowest possible entropy, zero in quantum mechanics, to minus infinity. Note that entropy is the logarithm for the count of states, which is quantified in units of $\hbar$. Therefore, taking $\hbar \to 0$ is equivalent to setting the entropy for pure states to minus infinity. This way, all finite entropies are large compared to the lowest possible entropy. In a nutshell, this is what we will show in section 4. Since physically it does not make sense to take limits of physical constants, $c \to \infty$ should be understood as $v \ll c$, $\hbar \to 0$ should be understood as $S \gg 0$.

@@

For a trapped condensate, the PSD at the center and entropy $S$ are related by $\text{PSD}=e^{5/2+\gamma-S/N}$, where $N$ is the number of particles and $\gamma$ is the virial coefficient for the trapping potential~\cite{PhysRevLett.78.990}. This has been used experimentally to condense at constant entropy, by changing the shape of the potential~\cite{PhysRevLett.81.2194}. Note that, however, the system is inhomogeneous, so the condensation happens because entropy is transferred from one part of the system to the other.

@@

Mathematically, the trace distance $T(\rho,\sigma)$ between two mixed states $\rho$ and $\sigma$ is contractive under a completely positive trace-preserving (CPTP) map $M$. That is, $T(M(\rho),M(\sigma)) \leq T(\rho,\sigma)$~\cite{nielsen2010quantum}. In the finite-dimensional case, we can have a CPTP map that increases the entropy of all states except that of the maximally mixed state, which is a fixed point.\footnote{This is a particular instance of mixing/relaxing ergodic quantum channels where the fixed point is the maximally mixed state~\cite{Burgarth2012ErgodicAM}.}  Under such a map, all pure states are mapped to mixed states, orthogonal states are mapped to non-orthogonal states and the map is strictly contractive. In section 4, when we construct the high-entropy limit, the map that results will follow this intuition in the infinite-dimensional case, with the only difference that there is no maximally mixed state that does not change.\footnote{In the infinite-dimensional case, one could have infinite-dimensional subspaces that remain orthogonal while increasing the entropy of all its elements. This corresponds, for example, to the case where there are conserved quantities. Note that in the map we construct in section 4, the expectation of all polynomials of position and momentum changes, and therefore no function of position and momentum is a conserved quantity. }  If we call $\mathcal{G}$ the space of all mixtures of Gaussian states, the states with a non-negative Wigner function, given an arbitrary mixed state $\rho$, we will find a state $\sigma_{\rho} \in \mathcal{G}$ that is closest to $\rho$. As we apply our entropy-increasing map, the entropy of $\rho$ increases and $\sigma_{\rho}$ will get closer to $\rho$. In this sense the states will ``look classical.''

@@

\section{The high-entropy limit}

@@

\bibliographystyle{plain}
\bibliography{bibliography}

>>>>

% =========================================================================
% SciPost LaTeX template
% Version 2024-07
%
% Submissions to SciPost Journals should make use of this template.
%
% INSTRUCTIONS: simply look for the `TODO:' tokens and adapt your file.
% ========================================================================

\documentclass{SciPost}

% Prevent all line breaks in inline equations.
\binoppenalty=10000
\relpenalty=10000

\hypersetup{
    colorlinks,
    linkcolor={red!50!black},
    citecolor={blue!50!black},
    urlcolor={blue!80!black}
}

\usepackage[bitstream-charter]{mathdesign}
\urlstyle{same}

%\usepackage[dvipsnames]{xcolor}  
%Had package options conflict.  Added option [dvipsnames] to SciPost.cls file
\usepackage{tikz-cd}
\usepackage{pgfplots}

% Fix \cal and \mathcal characters look (so it's not the same as \mathscr)
\DeclareSymbolFont{usualmathcal}{OMS}{cmsy}{m}{n}
\DeclareSymbolFontAlphabet{\mathcal}{usualmathcal}

\fancypagestyle{SPstyle}{
\fancyhf{}
\lhead{\colorbox{scipostblue}{\bf \color{white} ~SciPost Physics }}
\rhead{{\bf \color{scipostdeepblue} ~Submission }}
\renewcommand{\headrulewidth}{1pt}
\fancyfoot[C]{\textbf{\thepage}}
}

\begin{document}

\pagestyle{SPstyle}

\begin{center}{\Large \textbf{\color{scipostdeepblue}{
Classical mechanics as the high-entropy limit \\ of quantum mechanics}}}\end{center}

\begin{center}\textbf{
%%%%%%%%%% TODO: AUTHORS
% Write the author list here. 
% Use (full) first name (+ middle name initials) + surname format.
% Separate subsequent authors by a comma, omit comma and use "and" for the last author.
% Mark the corresponding author(s) with a superscript symbol in this order
% \star, \dagger, \ddagger, \circ, \S, \P, \parallel, ...
Gabriele Carcassi\textsuperscript{1},
Manuele Landini\textsuperscript{2} and
Christine A. Aidala\textsuperscript{1$\star$}
%%%%%%%%%% END TODO: AUTHORS
}\end{center}

\begin{center}
%%%%%%%%%% TODO: AFFILIATIONS
% Write all affiliations here.
% Format: institute, city, country
{\bf 1} Physics Department, University of Michigan, Ann Arbor, Michigan 48109, USA
\\
{\bf 2} Institut f\"ur Experimentalphysik und Zentrum f\"ur Quantenphysik, \protect\\ Universit\"at Innsbruck, Innsbruck, Austria
%%%%%%%%%% END TODO: AFFILIATIONS
%%%%%%%%%% TODO: EMAIL
% Provide email address of corresponding author(s)
\\[\baselineskip]
$\star$ \href{mailto:caidala@umich.edu}{\small caidala@umich.edu}

%%%%%%%%%% END TODO: EMAIL
\end{center}

\section*{\color{scipostdeepblue}{Abstract}}
\textbf{\boldmath{%
We show that classical mechanics can be recovered as the high-entropy limit of quantum mechanics. That is, the high entropy masks quantum effects, and mixed states of high enough entropy can be approximated with classical distributions. The mathematical limit $\hbar \to 0$ can be reinterpreted as setting the zero entropy of pure states to $-\infty$, in the same way that non-relativistic mechanics can be recovered mathematically with $c \to \infty$. Physically, these limits are more appropriately defined as $S \gg 0$ and $v \ll c$. Both limits can then be understood as approximations independently of what circumstances allow those approximations to be valid. Consequently, the limit presented is independent of possible underlying mechanisms and of what interpretation is chosen for both quantum states and entropy. 
}}

\vspace{\baselineskip}

%%%%%%%%%% BLOCK: Copyright information
% This block will be filled during the proof stage, and finilized just before publication.
% It exists here only as a placeholder, and should not be modified by authors.
\noindent\textcolor{white!90!black}{%
\fbox{\parbox{0.975\linewidth}{%
\textcolor{white!40!black}{\begin{tabular}{lr}%
  \begin{minipage}{0.6\textwidth}%
    {\small Copyright attribution to authors. \newline
    This work is a submission to SciPost Physics. \newline
    License information to appear upon publication. \newline
    Publication information to appear upon publication.}
  \end{minipage} & \begin{minipage}{0.4\textwidth}
    {\small Received Date \newline Accepted Date \newline Published Date}%
  \end{minipage}
\end{tabular}}
}}
}
%%%%%%%%%% BLOCK: Copyright information

%%%%%%%%%% TODO: LINENO
% For convenience during refereeing we turn on line numbers:
\linenumbers
% You should run LaTeX twice in order for the line numbers to appear.
%%%%%%%%%% END TODO: LINENO

%%%%%%%%%% TODO: TOC 
% Guideline: if your paper is longer that 6 pages, include a TOC
% To remove the TOC, simply cut the following block
\vspace{10pt}
\noindent\rule{\textwidth}{1pt}
\tableofcontents
\noindent\rule{\textwidth}{1pt}
\vspace{10pt}
%%%%%%%%%% END TODO: TOC

%%%%%%%%% TODO: CONTENTS 
% Write your article contents here, starting from first \section.
% An example structure is given below.

\section{Introduction}
\label{sec:intro}

@@

Mathematically, it is common to see both limits as group contractions. This too can be given a physically meaningful interpretation. The low-speed limit is mathematically recovered by taking the highest possible speed and setting it to infinity. This way, all finite velocities are small compared to the highest possible speed. This corresponds to taking $c \to \infty$. Conversely, the high-entropy limit is mathematically recovered by taking the lowest possible entropy, zero in quantum mechanics, to minus infinity. Note that entropy is the logarithm for the count of states, which is quantified in units of $\hbar$. Therefore, taking $\hbar \to 0$ is equivalent to setting the entropy for pure states to minus infinity. This way, all finite entropies are large compared to the lowest possible entropy. In a nutshell, this is what we will show in section~\ref{sec:high_entropy_limit}. Since physically it does not make sense to take limits of physical constants, $c \to \infty$ should be understood as $v \ll c$, $\hbar \to 0$ should be understood as $S \gg 0$.

@@

For a trapped condensate, the PSD at the center and entropy $S$ are related by $\text{PSD}=$ $e^{5/2+\gamma-S/N}$, where $N$ is the number of particles and $\gamma$ is the virial coefficient for the trapping potential~\cite{PhysRevLett.78.990}. This has been used experimentally to condense at constant entropy, by changing the shape of the potential~\cite{PhysRevLett.81.2194}. Note that, however, the system is inhomogeneous, so the condensation happens because entropy is transferred from one part of the system to the other.

@@

Mathematically, the trace distance $T(\rho,\sigma)$ between two mixed states $\rho$ and $\sigma$ is contractive under a completely positive trace-preserving (CPTP) map $M$. That is, $T(M(\rho),M(\sigma)) \leq$ $T(\rho,\sigma)$~\cite{nielsen2010quantum}. In the finite-dimensional case, we can have a CPTP map that increases the entropy of all states except that of the maximally mixed state, which is a fixed point.\footnote{This is a particular instance of mixing/relaxing ergodic quantum channels where the fixed point is the maximally mixed state~\cite{Burgarth2012ErgodicAM}.}  Under such a map, all pure states are mapped to mixed states, orthogonal states are mapped to non-orthogonal states and the map is strictly contractive. In section~\ref{sec:high_entropy_limit}, when we construct the high-entropy limit, the map that results will follow this intuition in the infinite-dimensional case, with the only difference that there is no maximally mixed state that does not change.\footnote{In the infinite-dimensional case, one could have infinite-dimensional subspaces that remain orthogonal while increasing the entropy of all its elements. This corresponds, for example, to the case where there are conserved quantities. Note that in the map we construct in section 4, the expectation of all polynomials of position and momentum changes, and therefore no function of position and momentum is a conserved quantity. }  If we call $\mathcal{G}$ the space of all mixtures of Gaussian states, the states with a non-negative Wigner function, given an arbitrary mixed state $\rho$, we will find a state $\sigma_{\rho} \in \mathcal{G}$ that is closest to $\rho$. As we apply our entropy-increasing map, the entropy of $\rho$ increases and $\sigma_{\rho}$ will get closer to $\rho$. In this sense the states will ``look classical.''

@@

\section{The high-entropy limit}
\label{sec:high_entropy_limit}

@@

\section*{Acknowledgements}
This paper is part of the ongoing Assumptions of Physics project~\cite{aop-book}, which aims to identify a handful of physical principles from which the basic laws can be rigorously derived.

% TODO: include funding information
\paragraph{Funding information}
ML acknowledges funding via the FWF-funded QuantERA project QuSiED with project number I 6008-N.

%%%%%%%%% END TODO: CONTENTS

% Use your bibtex library, formatted by the SciPost style file.
\bibliography{bibliography.bib}

%%%%%%%%%% END TODO: BIBLIOGRAPHY
